Synthesizing Quotient Lenses

### 2017-11-10 papers/QuotientLensSynth/main.tex
Made changes

\section{Introduction}
A {\em lens} $\ell$ is a {\em bidirectional transformation} between a source
type $S$ and a target type $T$ in which the forward transformation
$\ell.\get$ and the backward transformation $\ell.\lput$ satisfy certain
round-tripping laws such as invertibility. Lenses are advantageous
because writing a single lens rather than two separate uni-directional
transformations saves both keystrokes and bugs, by reducing redundancy and by
automatically guaranteeing strong ``lens laws'' ensuring that the forward and
backward transformations interact well.

@@

This paper addresses both of these issues.  First, in Section~\ref{sec:example} we
explain the difficulties of programming with quotient lenses and show, by way of
example, how our novel language of {\em Quotient Regular Expressions}, or QREs
and QRE lenses address these difficulties.  Quotient regular expressions allow
programmers to specify a broad class of equivalence relations over regular
languages using a compact and convenient notation in which the regular language
and its equivalence classes are captured by a single expression.  We then
define {\em QRE lenses}, a class of bijective quotient lenses that are typed
using QREs, and we show how to {\em synthesize} QRE lenses from their QRE types
and a set of input-output examples.  We demonstrate the practicality of this
approach by synthesizing QRE lenses between several real-world data formats
from the {\tt data.gov} database.

\saz{There is still some redundancy between the following contributions list and
  the paragraph above.}

In summary, our main contributions are:
\begin{enumerate}
\item We introduce a novel language of {\em Quotient Regular Expressions}
(QREs) that provide compact, convenient notation for a natural and useful
class of equivalence relations on regular languages (Section~\ref{QRE}).
\item We define a language of {\em QRE lenses}, a subset of bijective quotient
lenses whose types are given by QREs. Our main technical contribution in this
paper is a proof that each QRE lens $q$ defines the same transformation as a
QRE lens $q'$ that is in normal form. (Section~\ref{QRE-lenses}).
\item Using this normal form, we reduce the problem of {\em synthesizing}
QRE lenses from their QRE types and a set of examples to the previously studied
problem of synthesizing bijective lenses between regular languages 
(Section~\ref{synth}).
\item We extend the Boomerang {\em implementation} with QREs, QRE lens operators,
and QRE lens synthesis and demonstrate its applicability by using it to
synthesize QRE lenses between several real-world data formats from the
{\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

\section{Quotient Regular Expressions and Lenses by Example}
\label{sec:example}

@@

We wish for this transformation to have the same behaviour on records that
differ only in ``inessential'' details. First, the transformation
should allow whitespace characters to occur arbitrarily between the words in the
label, author and title fields as well as in between them. Second, an
author's name can have the surname followed by a comma and then the other
names separated by arbitrary whitespace, or just all the names separated by
arbitrary whitespace (for example, the names ``Conway, J. H.'' and ``J. H.
Conway'' are equivalent). Lastly, the transformation should also ignore the
order in which the author and title fields occur in each citation record.

@@

While it is possible to express the equivalence on the \bibtex{} data above in
Boomerang, actually doing so is a challenge. This is because the entire
equivalence is composed of three separate equivalences \saz{explicitly mention
  whitespace and permutation?} which nest and interact
with each other in a way that makes them difficult to express with the standard
Boomerang combinators. Moreover, in many cases, Boomerang requires that the
programmer first specify the data itself in the form of a regular expression
before she can define the equivalence relation on that data. This leads to
verbose, repetitive code.

Instead, we propose {\em Quotient Regular Expressions} or QREs which enable
the programmer to define a regular expression $R$ as well as an equivalence
relation on $R$ in the same expression.For example, with QREs we can specify
\bibtex{} records as shown in Figure~\ref{fig:example-qre}.

@@

(* project QRE maps arbitrary occurrences of whitespace to a single space *)
let wsp_to_space   = project wsp$^+$ $\to$ " "

@@

(* perm QRE allows permutations of fields with arbitrary whitespace in between *)
let bib_perm  = perm (label, bib_author, bib_title) with (project ("," .wsp$^+$) $\to$ ",\n")
let bibtex    = "@book{" . bib_perm . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
  \label{fig:example-qre}
\end{figure}

In Figure~\ref{fig:example-qre}, the QRE \cd{wsp_to_space}, which is defined to
be

\noindent \cd{project wsp}+ $\to$ \cd{" "}, matches any non-zero number of
whitespace characters, but considers the matched string of whitespace
characters as equivalent with a single space as the canonical representative. 
We call this QRE a \textit{projection} because it projects an
entire regular language down to a single representative string in the
language.

Another form of QRE is the \textit{squash} QRE, an example of which is given by
\cd{valid_name}, which is defined to be:

@@

\noindent This QRE specifies that the \cd{comma_name} representation of the
author's name is equivalent to the \cd{space_name} representation.  Here \cd{get
  space_to_comma} is the function that transforms names in the \cd{comma_name}
representation to names in the \cd{space_name} representation.  In this
  example, \cd{space_to_comma} is itself an automatically \textit{synthesized}
  lens between the \cd{space_name} and \cd{comma_name} QREs, and we need to use
  the ``get'' direction of the lens to specify the squash transformation. (We'll
  say more about synthesis below.)

The third example of a QRE in the Figure is \cd{bib_perm}, defined as:
\begin{center}
\begin{lstlisting}
perm (label, bib_author, bib_title) with (project ("," . wsp$^+$ ) $\to$ ",\n")
\end{lstlisting}
\end{center}

\noindent This is a \textit{permutation} QRE, which specifies that the label,
title, and author fields can occur in any order, with the fields separated by a
comma followed by arbitrary amounts of whitespace. The canonical representation
is that in which the label, author and title fields occur in that order,
separated by a comma followed a newline.
Such \cd{perm} QREs allow for easy nesting of permutations. Similarly, it is easy
to specify any number of fields within the \cd{perm} QRE.

Given the \textit{projection}, \textit{squash}, and \textit{permutation}
  QRE primitives, we can build more complex languages using the usual regular
  expression combinators of alternative choice `\cd{|}', concatenation `\cd{.}'
  and Kleene star `$^*$', and also sequential composition ``;''.  As shown in
  the rest of Figure~\ref{fig:example-qre}, it is straightforward to define both a
  regular language that accepts \bibtex{} records \textit{and}, simultaneously,
  an equivalence relation that quotients out irrelevant details.

\subsection{QRE Lenses}
QREs make it easier to write quotient lenses by specifying the data and the
equivalence relation on the data in one expression, but to use them to define a
lens, the programmer still has to manually derive the underlying lens that maps
the equivalence classes of the source data to the equivalence classes of the
target data.

@@

This is itself a tedious and potentially error-prone task, as the code snippet
in Figure~\ref{fig:example-lens} illustrates.  The lens shown in the Figure
transforms \bibtex{} data to EndNote data  (and vice versa) by matching
the delimiters and separators in a bijective manner. For example, the
\cd{preamble} lens replaces the \bibtex{} preamble \verb|@Book| with the EndNote
preamble \verb|%0 Book\n%F|. When the programmer is writing the lens, she has
to make sure that all of these small deletions and insertions are correct,
which can be a challenging task if the lens is large enough. This makes lens
programming tedious and error-prone.

To address this challenge, we identify a  class of quotient lenses that we call
{\em QRE lenses}, and show how to {\em synthesize} QRE lenses from a pair of
QREs and a set of input-output example pairs. Our key insight is to reduce the
problem of synthesizing QRE lenses to the problem of synthesizing {\em
bijective lenses}, which are the bijections created from the Boomerang lens
combinators. This is a problem which we solved in our prior work ~\cite{popl18}.

@@

For example, suppose \cd{q_1 : bibtex} $\Leftrightarrow$ \cd{endnote} is a QRE
lens that maps between \bibtex{} and EndNote formats, and that \cd{q_2 : endnote
  <=> xml} is a QRE lens that maps between EndNote and XML formats. The
composition \cd{(q_1 ; q2) : bibtex} $\Leftrightarrow$ \cd{xml} of \cd{q_1} and
\cd{q_2} is not (necessarily) in normal form due to the presence of canonizers
on each side of the intermediary EndNote format:

@@

\noindent However, while \cd{q_1 ; q_2} is not in normal form, our main theorem
shows that there is another QRE lens \cd{q : bibtex} $\Leftrightarrow$ \cd{xml}
of the same type as \cd{q_1 ; q_2} such that \cd{q_1 ; q_2} and \cd{q} define
the same transformation. This proof will allow justify our approach for
synthesizing quotient lenses so that the programmer can replace all of the
code in Figure~\ref{fig:example-lens} with a call to the synthesis prodedure:

@@

\subsection{Syntax of Quotient Regular Expressions}
The language of Quotient Regular Expressions (QREs) is given by the following
grammar: 
\begin{align*}
c := \; &R \sep \project{R}{s} \sep \squash{c}{c'}{f} \sep
\perm{c_1, \ldots, c_n}{c} \\
  & | \; \normalize{R}{R'}{f} \sep c \semicolon c' \sep c \cdot c' \sep (c \sep
  c') \sep c^*,
\end{align*}
where $c$ ranges over QREs, $R$ ranges over regular expressions, $f$ ranges over
functions between regular languages, and $s$ ranges over character strings. In
subsequent sections, we will use the the notation $\mathcal{L}(R)$ to mean the
language accepted by the regular expression $R$.

@@

\begin{lstlisting}
(* Regular expressions matching different representations *)
let comma_name = last_name . "," . wsp+ . first_name
let space_name = first_name . wsp+ . last_name

(* Squash comma representation to space representation *)
let comma_to_space = 
synth comma_name <=> space_name using {("Conway, J. H.", "J. H. Conway")} 
let comma_to_space = squash comma_name $\to$ space_name using (get comma_to_space)

(* project whitespace in space_name representation *)
let project_in_space = first_name . (project wsp+ $\to$ " ") . last_name

@@

The equivalence relations definable by QREs are a strict subset of those that
can be expressed in Boomerang. For instance, in Boomerang, one can map a
regular language described by $R$ to a regular language described by $S$, with
$\mathcal{L}(S) \not \subseteq \mathcal{L}(R)$, whereas each QRE $c$ canonizes
a regular language described by $R$ to regular subset $\mathcal{L}(R')
\subseteq \mathcal{L}(R)$ described by $R'$.

@@

Figure~\ref{fig:qrerules} gives the inferrence rules for derving well-formed
QREs. As expected, the unambiguity conditions come into play when defining QREs
using the regular combinators. For example, the inferrence rule for
concatenation is as follows:
\begin{prooftree}
\AxiomC{$c$ \textbf{WF}}
\AxiomC{$c'$ \textbf{WF}}
\AxiomC{$W(c) \cdot^! W(c')$}
\AxiomC{$K(c) \cdot^! K(c')$}
\QuaternaryInfC{$c \cdot c'$ \textbf{WF}}
\end{prooftree}

\noindent This rule says that we can concatenate two QREs $c$ and $c'$ only when
the concatentions of $W(c)$ and $W(c')$, and $K(c)$ and $K(c')$ are ambiguous.

@@

The third hypothesis says that the regular expression $K(c_1) \cdot K(c)
\cdot \ldots \cdot K(c) \cdot K(c_n)$ must be unambiguous. This is because the
underlying lens that maps to or from a \textit{permutation} QRE operates on
the language $K(c_1) \cdot K(c) \cdot \ldots \cdot K(c) \cdot K(c_n)$ which is
the kernel of the \textit{permutation} QRE, and as we mentioned, the underlying
lens will require that the source and target regular expressions of the lens be
unambiguous so that the lens can match strings uniquely.

@@

As we mentioned before, each QRE $c$ enables us to express:
\begin{enumerate}
  \item a regular expression $W(c)$ (the ``whole'' of $c$),
  \item an equivalence relation $\eqrel{c}$ on $\mathcal{L}(W(c))$,
  \item a regular expression $K(c)$ (the ``kernel'' of $c$)
  such that $\mathcal{L}(K(c))$ forms a complete set of representatives for
  $\eqrel{c}$, and
  \item a ``canonizing'' function $\canonizer(c):\mathcal{L}(W(c))
  \longrightarrow \mathcal{L}(K(c))$ which given any $w \in \mathcal{L}(W(c))$,
  computes $\canonizer(c)(w)$ as the unique $k$ in $\mathcal{L}(K(c))$ such that
  $k$ is equivalent to $w$ mod $\eqrel{c}$
  \end{enumerate}
Figures~\ref{fig:wk} gives the inductive rules that define the whole language
$W(c)$ of a QRE. The two interesting rules are the rules for the
\textit{squash}, \textit{normalize} and \textit{permutation} combinators. If $c
= \squash{c}{c'}{f}$, then the whole language of $c$ is $W(c) \sep W(c')$. This
is because the \textit{squash} combinator takes the whole language $W(c)$ of
$c$, the whole language $W(c')$ of $c'$ and a function $f : \mathcal{L}(W(c))
\longrightarrow \mathcal{L}(W(c'))$, maps $\mathcal{L}(W(c))$ into
$\mathcal{L}(W(c'))$ using $f$ and then canonizes $W(c')$ into $K(c')$ using
$\canonizer(c')$.

For the $\perm{(c_1, \ldots, c_n)}{c}$ combinator, the whole language is the
union of languages of the form $W(c_{\sigma(1)}) \cdot W(c) \cdot \ldots \cdot
W(c) \cdot W(c_{\sigma(n)})$. This is because the \textit{permutation}
combinator allows for the string to match any permutation the $c_i$'s while also
taking into account the separator $c$ in between each of the $c_i$'s. The kernel
of the \textit{permutation} combinator is the language $K(c_1) \cdot K(c) \cdot
\ldots \cdot K(c) \cdot K(c_n)$, because the canonical permutation is the
identity permuation, and then each of the parts of the string that match $c_i$
and $c$ are canonized into $K(c_i)$ and $K(c)$ respectively.

Figure~\ref{fig:canonizers} gives the inductive definitions of the
$\canonizer{}$ function for each QRE. Again, the the least obvious definition
is

@@

Our approach in defining QRE lenses is as follows. Let $c$ and $c'$ be QREs. Given a
bijection $\ell : K(c) \Leftrightarrow K(c')$, we may define a quotient lens $q
: W(c)/\eqrel{c} \Leftrightarrow W(c')/\eqrel{c'}$ by
\begin{equation}\label{normalform}
q.get = \ell \circ \canonizer(c) \text{ and } q.put = \ell^{-1} \circ
\canonizer(c'),
\end{equation}

\noindent In other words, the get function of the QRE first canonizes the source
data using $\canonizer(c)$ and then applies $f$ to the result, while the
put function of the QRE first canonizes the target data using
$\canonizer(c')$ and the applies $f^{-1}$ to the result.

One potential pitfall with QRE lenses is that the combinators that
we define on them may not preserve this normal form. For example, the
composition $q' \; ; \; q$ of QRE lenses $q$ and $q'$ may not be equivalent to
any quotient lens that has the canonical normal form of a bijective lens with
canonizers a the edges. Consequently, in order to define quotient lenses, we
need to choose a set $S$ of bijections that has the right properties in the
sense that QRE lenses built from the bijections in $S$ are closed under the
action of QRE lens combinators. Our choice is to use the class of
\textit{bijective lenses}, which are the bijections that can be expressed in
Boomerang. In the next subsection we briefly introduce bijective lenses and
discuss the properties of bijective lenses that we will use to prove that QRE
lenses have the normal form of bijective lens with canonizers at the edges.

@@

The more interesting combinators are the $\kw{lquot}$ and $\kw{rquot}$
combinators. The $\kw{lquot}(c, q)$ combinator takes a quotient lens $q$ and
quotients the source data using $c$, assuming that the source data forms a
complete set of representatives for the equivalence relation $\eqrel{c}$. The
$\kw{lquot}(q, c)$ combinator does the same but on the view data.

@@

The denotation $\llbracket q \rrbracket$ of a QRE lens $q:c \Leftrightarrow c'$
is a quotient lens $\llbracket q \rrbracket : W(c)/{\eqrel{c}}
\Longleftrightarrow W(c')/{\eqrel{c'}}$. The typing rules and denotation of QRE
lenses are given in Figure~\ref{fig:qlenssemantics}.

The following theorem confirms that the semantics for QRE lenses are correct:
\begin{theorem}
If there is a derivation $q : c \Leftrightarrow c'$, then $\llbracket q
\rrbracket : W(c)/{\eqrel{c}} \Leftrightarrow W(c')/{\eqrel{c'}}$ is a
well-defined quotient lens.
\end{theorem}

>>>>

\section{Introduction}
{\em Lenses}~\cite{Lenses} specify bidirectional 
transformations between different data formats.  Writing a single lens
rather than two separate uni-directional transformations saves both
keystrokes and bugs, by reducing redundancy and by automatically
guaranteeing strong ``lens laws'' ensuring that the forward and backward
transformations interact well.

@@

In this paper, we make programming quotient lenses easier in two ways. First, we
introduce a novel language of {\em Quotient Regular Expressions}, or QREs, which
enable a programmer to simultaneously specify a regular expression as well as an
equivalence relation on that regular expression. QREs make programming quotient
lenses easire by saving the programmer the effort of defining the data and the
equivalence relation on the data separately.

Secondly, we use QREs to define a class of quotient lenses called {\em QRE
lenses} which map betweeen QREs, and show how to synthesize QRE lenses given
an specification described by pair of QREs and a set of input-output examples.
This way, the programmer need not manually derive the lens herself. The net
effect of these two contributions, that is QREs and QRE lens synthesis, is that
in order to derive a quotient lens, the programmer will only need to specify the
data formats on each side of the desired quotient lens and possibly a set of
example input-output pairs that guide synthesis of the lens, a marked
improvement over the current situation.

In summary, our main contributions are:
\begin{enumerate}
\item We introduce a novel language of {\em Quotient Regular Expressions}
(QREs) that a provide compact, convenient notation for a natural and useful
class of equivalence relations on regular languages (Section~\ref{QRE}).
\item We define a language of {\em QRE lenses}, a subset of bijective quotient
lenses whose types are given by QREs. Our main technical contribution in this
paper is a proof that each QRE lens $q$ defines the same transformation as a
QRE lens $q'$ that is in normal form. (Section~\ref{QRE-lenses}).
\item Using this normal form, we reduce the problem of {\em synthesizing}
QRE lenses from their QRE types and a set of examples to the previously studied
problem of synthesizing bijective lenses between regular languages 
(Section~\ref{synth}).
\item We extend the Boomerang {\em implementation} with QREs, QRE lens operators,
and QRE lens synthesis and demonstrate its applicability by using it to
synthesize QRE lenses between several real-world data formats from the
{\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

\section{Quotient Regular Expressions and Lenses by Example}
\label{sec:example}
In this section, we explain the difficulties of programming with quotient lenses
and show, by way of example, how our novel language of {\em Quotient Regular
Expressions}, or QREs and QRE lenses address these difficulties.

@@

We wish for this transformation to have the same behaviour on records that
differ only in ``inessential'' details. First, the transformation
should allow whitespace characters to occur arbitrarily between the words in the
label, author and title fields as well, as in between them. Second, an
author's name can have the surname followed by a comma and then the other
names separated by arbitrary whitespace, or just all the names separated by
arbitrary whitespace (for example, the names ``Conway, J. H.'' and ``J. H.
Conway'' are equivalent). Lastly, the transformation should also ignore the
order in which the author and title fields occur in each citation record.

@@

While it is possible to express the equivalence on the \bibtex{} data above in
Boomerang, actually doing so is a challenge. This is because the entire
equivalence is composed of three separate equivalences---whitespace quotienting
inside fields, whitespace quotenting in between fields, and permutation
quotenting of fields---nest with each other in a way that makes them difficult
to express with the standard Boomerang combinators.

In general, there is also another issue in describing equivalences in
Boomerang, which is that that the programmer often has to first specify the data
itself in the form of a regular expression before she can define the
equivalence relation on that data. This leads to repetitive code since the
regular expression that defines the data and the equivalence relation defined on
the regular expression follow the same syntactic patter---one can do both things
with the same code with only a few changes.

\sam{TODO: PUT EXAMPLE HERE!}

To address these issues, we propose {\em Quotient Regular Expressions} or QREs
which enable the programmer to define a regular expression $R$ as well as an
equivalence relation on $R$ in the same expression. For example, with QREs we
can specify \bibtex{} records as shown in Figure~\ref{fig:example-qre}.

@@

(* map arbitrary occurrences of whitespace to a single space *)
let wsp_to_space   = project wsp$^+$ $\to$ " "

@@

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_perm  = perm (label, bib_author, bib_title) with (project ("," .wsp$^+$) $\to$ ",\n")
let bibtex    = "@book{" . bib_perm . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
  \label{fig:example-qre}
\end{figure}

In Figure~\ref{fig:example-qre}, the QRE \cd{project wsp}+ $\to$ \cd{" "}
matches any non-zero number of whitespace characters, but considers the matched
string of whitespace characters as equivalent with a single space as the
canonical representative. We call this QRE a \textit{projection} because it
projects an entire regular language down to a single representative string in
the language.

Another QRE in Figure~\ref{fig:example-qre} is given by

@@

\noindent This QRE is a \textit{squash} QRE. A \textit{squash} QRE ``glues''
two data formats using a user-defined function. For example,
the QRE in the code above specifies that the \cd{comma_name} representation of
the author's name is equivalent to the \cd{space_name} representation, with
\cd{space_name} taken to be the canonical representation. The
function that identifies the two formats is given by \cd{get space_to_comma},
which transforms names in the \cd{comma_name} representation to names in the
\cd{space_name} representation. In this example, \cd{space_to_comma} is itself
an automatically \textit{synthesized} lens between the \cd{space_name} and
\cd{comma_name} QREs, and we need to use the ``get'' direction of the lens to
specify the squash transformation. (We'll say more about synthesis below.)

The third example of a QRE in the Figure is \cd{bib_perm}, defined by:
\begin{center}
\begin{lstlisting}
perm (label, bib_author, bib_title) with (project ("," . wsp$^+$ ) $\to$ ",\n")
\end{lstlisting}
\end{center}

\noindent This is a \textit{permutation} QRE that specifies that the label,
title, and author fields can occur in any order, with the fields separated by a
comma followed by arbitrary amounts of whitespace. The canonical representation
is that in which the label, author and title fields occur in that order,
separated by a comma followed a newline. The \textit{permutation} QREs is
designed for equivalences such as this, where different permutations of
equivalences are interspersed with a separator which itself has an
equivalence defined on it.

Given the \textit{projection}, \textit{squash}, and \textit{permutation}
  QRE primitives, we can build more complex languages using the usual regular
  expression combinators of alternative choice `\cd{|}', concatenation `\cd{.}'
  and Kleene star `$^*$', and also sequential composition ``;''.

\subsection{QRE Lenses}
While QREs make it easier to write equivalence relations on regular expressions
by specifying the data and the equivalence relation on the data in one
expression, in order to define a quotient lens, the programmer still has to
manually derive the underlying lens that maps the equivalence classes of the
source data to the equivalence classes of the target data. This is itself a
tricky task as Figure~\ref{fig:example-lens} illustrates.

@@

The lens shown in Figure~\ref{fig:example-lens} transforms \bibtex{} data to
EndNote data  (and vice versa) by matching the delimiters and separators in a
bijective manner. For example, the \cd{preamble} lens replaces the \bibtex{}
preamble \verb|@Book| with the EndNote preamble \verb|%0 Book\n%F|. When the
programmer is writing the lens, she has to make sure that all of these small
deletions and insertions are correct, which can be a challenging task if the
lens is large enough, as in the definition of the lens \verb!author_lens! in
Figure~\ref{fig:example-lens}. This makes lens programming tedious and
error-prone.

Consequently, we offer the programmer some relief by identifying a class of
quotient lenses that we call {\em QRE lenses}, and show how to {\em synthesize}
QRE lenses from a pair of QREs and a set of input-output example pairs. Our key
insight is to reduce the problem of synthesizing QRE lenses to the problem of
synthesizing {\em bijective lenses}, which are the bijections created from the
Boomerang lens combinators. This is a problem which we solved in our prior work
~\cite{popl18}.

@@

For example, suppose \cd{q_1 : bibtex} $\Leftrightarrow$ \cd{endnote} is a QRE
lens that maps between \bibtex{} and EndNote formats, and that \cd{q_2 : endnote
  <=> xml} is a QRE lens that maps between EndNote and XML formats. The
composition \cd{(q_1 ; q2) : bibtex} $\Leftrightarrow$ \cd{xml} of \cd{q_1} and
\cd{q_2} is not necessarily in normal form due to the presence of canonizers
on each side of the intermediary EndNote format:

@@

\noindent However, while \cd{q_1 ; q_2} is not in normal form, our main theorem
shows that there is another QRE lens \cd{q : bibtex} $\Leftrightarrow$ \cd{xml}
of the same type as \cd{(q_1 ; q2) : bibtex} $\Leftrightarrow$ \cd{xml} such
that \cd{q_1 ; q_2} and \cd{q} define the same transformation. This proof
will allow justify our approach for synthesizing quotient lenses so that the
programmer can replace all of the code in Figure~\ref{fig:example-lens} with a
call to the synthesis prodedure:

@@

\subsection{Syntax of Quotient Regular Expressions}
The language of Quotient Regular Expressions (QREs) is given by the following
grammar: 
\begin{align*}
c := \; &R \sep \project{R}{s} \sep \squash{c}{c'}{f} \sep
\perm{c_1, \ldots, c_n}{c} \\
  & | \; \normalize{R}{R'}{f} \sep c \semicolon c' \sep c \cdot c' \sep (c \sep
  c') \sep c^*,
\end{align*}
where $c$ ranges over QREs, $R$ ranges over regular expressions, $f$ ranges over
functions between regular languages, and $s$ ranges over character strings. In
the following sections, we will use the the notation $\mathcal{L}(R)$ to mean
the language accepted by the regular expression $R$.

@@

\begin{lstlisting}
(* Regular expressions matching different comma and space representations *)
let comma_name = last_name . "," . wsp+ . first_name
let space_name = first_name . wsp+ . last_name

(* Squash comma representation into space representation *)
let comma_to_space = 
synth comma_name <=> space_name using {("Conway, J. H.", "J. H. Conway")} 
let comma_to_space = squash comma_name $\to$ space_name using (get comma_to_space)

(* Project away whitespace in space_name representation *)
let project_in_space = first_name . (project wsp+ $\to$ " ") . last_name

@@

The equivalence relations definable by QREs are a strict subset of those that
can be expressed in Boomerang. For instance, in Boomerang, one can use any lens
to canonize a regular language described by $R$ to a regular language described
by $S$, with $\mathcal{L}(S) \not \subseteq \mathcal{L}(R)$, whereas each QRE
$c$ canonizes a regular language described by $R$ to regular subset
$\mathcal{L}(R') \subseteq \mathcal{L}(R)$ described by $R'$. QREs therefore are
not as expressive as Boomerang canonizers. On the other hand, QREs enable us to
express many equivalences which occur in real world data as we shall demonstrate
in the evaluation section (Section~\ref{impl}).

@@

Figure~\ref{fig:qrerules} gives the inferrence rules for deriving well-formed
QREs. As expected, the unambiguity conditions come into play when defining QREs
using the regular combinators. For example, the inferrence rule for
concatenation is as follows:
\begin{prooftree}
\AxiomC{$c$ \textbf{WF}}
\AxiomC{$c'$ \textbf{WF}}
\AxiomC{$W(c) \cdot^! W(c')$}
\AxiomC{$K(c) \cdot^! K(c')$}
\QuaternaryInfC{$c \cdot c'$ \textbf{WF}}
\end{prooftree}

\noindent This rule says that the concatenation $c \cdot c'$ of QREs $c$ and
$c'$ is well formed only if the concatentions of $W(c)$ and $W(c')$, and
$K(c)$ and $K(c')$ are ambiguous.

@@

The third hypothesis says that the regular expression $K(c_1) \cdot K(c)
\cdot \ldots \cdot K(c) \cdot K(c_n)$ must be unambiguous. This is because the
underlying lens that maps to or from a \textit{permutation} QRE operates on
the language $K(c_1) \cdot K(c) \cdot \ldots \cdot K(c) \cdot K(c_n)$ which is
the kernel of the \textit{permutation} QRE. As we mentioned, the underlying
lens will require that the source and target regular expressions of the lens be
unambiguous so that the lens can match strings uniquely.

@@

As we mentioned before, each QRE $c$ enables us to express:
\begin{enumerate}
  \item a regular expression $W(c)$ (the ``whole'' of $c$),
  \item an equivalence relation $\eqrel{c}$ on $\mathcal{L}(W(c))$,
  \item a regular expression $K(c)$ (the ``kernel'' of $c$)
  such that $\mathcal{L}(K(c))$ forms a complete set of representatives for
  $\eqrel{c}$, and
  \item a ``canonizing'' function $\canonizer(c):\mathcal{L}(W(c))
  \longrightarrow \mathcal{L}(K(c))$ which given any $w \in \mathcal{L}(W(c))$,
  computes $\canonizer(c)(w)$ as the unique $k$ in $\mathcal{L}(K(c))$ such that
  $k$ is equivalent to $w$ mod $\eqrel{c}$
  \end{enumerate}
Figures~\ref{fig:wk} gives the inductive definitions for the whole language
$W(c)$ of a QRE. The \textit{squash} and \textit{permutation} combinators
give the two interesting definitions . If $c = \squash{c}{c'}{f}$, then
the whole language of $c$ is $W(c) \sep W(c')$. This is because the
\textit{squash} combinator takes the whole language $W(c)$ of $c$, the whole
language $W(c')$ of $c'$ and a function $f : \mathcal{L}(W(c)) \longrightarrow
\mathcal{L}(W(c'))$, maps $\mathcal{L}(W(c))$ into $\mathcal{L}(W(c'))$ using
$f$ and then canonizes $W(c')$ into $K(c')$ using $\canonizer(c')$.

For the $\perm{(c_1, \ldots, c_n)}{c}$ combinator, the whole language is the
union of languages of the form $W(c_{\sigma(1)}) \cdot W(c) \cdot \ldots \cdot
W(c) \cdot W(c_{\sigma(n)})$. This is because the \textit{permutation}
combinator allows for the string to match any permutation the $c_i$'s while also
taking into account the separator $c$ in between each of the $c_i$'s. The kernel
of the \textit{permutation} combinator is the language $K(c_1) \cdot K(c) \cdot
\ldots \cdot K(c) \cdot K(c_n)$, because the canonical permutation is the
identity permuation, with each of the parts of the string that match $c_i$
and $c$ canonized into $K(c_i)$ and $K(c)$ respectively.

Figure~\ref{fig:canonizers} gives the inductive definitions of the
$\canonizer{}$ function for each QRE. Again, the \textit{permutation} combinator
gives the most interesting definition,

@@

Given regular expressions $R, S$ and equivalence relations
$\sim_R$ and $\sim_S$ defined on $\mathcal{L}(R)$ and $\mathcal{L}(S)$
respectively, we define a {\em bijective quotient lens} $q : R /{\sim_R}
\Leftrightarrow S/{\sim_S}$ to be a pair of functions $q.get :
\mathcal{L}(R) \longrightarrow \mathcal{L}(S)$ and $q.put : \mathcal{L}(S)
\longrightarrow \mathcal{L}(R)$ such that for all $r \in \mathcal{L}(R)$ and $s
\in \mathcal{L}(S)$,
\begin{align*}
q.put(q.get(r)) &\sim_R r \text{, and}\\
q.get(q.put(s)) &\sim_R s
\end{align*}
Moreover, if $r \sim_R r'$, then $q.get(r) = q.get(r')$, and if $s \sim_S s'$,
then $q.put(s) = q.put(s')$.

In words, a bijective quotient lens $q$ from $R$ to $S$ modulo $\sim_R$
and $\sim_S$ is a pair of functions $q.get$ and $q.put$ such that $q.get$
respects $\sim_R$ and $q.put$ respects $\sim_S$, and such that the respective
functions defined on $\mathcal{L}(R)/{\sim_R}$ and $\mathcal{L}(S)/{\sim_S}$ are
mutual inverses of each other.

Now let $c$ and $c'$ be QREs. Given a bijection $\ell : K(c) \Leftrightarrow
K(c')$, we may define a quotient lens $q : W(c)/\eqrel{c} \Leftrightarrow
W(c')/\eqrel{c'}$ by
\begin{equation}\label{normalform}
q.get = \ell \circ \canonizer(c) \text{ and } q.put = \ell^{-1} \circ
\canonizer(c'),
\end{equation}

\noindent In other words, the get function of the $q$ first canonizes the source
data using $\canonizer(c)$ and then applies $f$ to the result, while the
put function of the $q$ first canonizes the target data using
$\canonizer(c')$ and the applies $f^{-1}$ to the result. This is our approach in
defining QRE lenses.

One potential pitfall with QRE lenses is that the combinators that
we define on them may not preserve this normal form. For example, the
composition $q' \; ; \; q$ of QRE lenses $q$ and $q'$ may not be equivalent to
any quotient lens that has the canonical normal form of a bijective lens with
canonizers a the edges. Consequently, in order to guarantee that every QRE
lens $q$ is equivalent to a QRE lens $q'$ which has the canonical form of
a bijection with canonizers at the edges, we need to identify a set $S$ of
bijections that will ensure that QRE lenses are closed under the action of QRE
lens combinators. Our choice is to use the class of \textit{bijective lenses},
which we briefly discuss in the next section.

@@

The interesting combinators are the $\kw{lquot}$ and $\kw{rquot}$
combinators. The $\kw{lquot}(c, q)$ combinator takes a quotient lens $q$ and
quotients the source data using $c$, assuming that the source data forms a
complete set of representatives for the equivalence relation $\eqrel{c}$. The
$\kw{lquot}(q, c)$ combinator does the same but on the target data.

@@

The denotation $\llbracket q \rrbracket$ of a QRE lens $q:c \Leftrightarrow c'$
is a quotient lens $\llbracket q \rrbracket : W(c)/{\eqrel{c}}
\Longleftrightarrow W(c')/{\eqrel{c'}}$. The typing rules and denotation of QRE
lenses are given in Figure~\ref{fig:qlenssemantics}. The trickiest typing rule
is the typing rule for composition:

  \begin{prooftree}
\AxiomC{$\substack{q_1 : c \Leftrightarrow c' \\ q_2 : c'' \Leftrightarrow
c'''}$}
\AxiomC{$\substack{\mathcal{L}(W(c'))= \mathcal{L}(W(c'')) \\ K(c') \equiv
K(c'')}$}
\AxiomC{$\canonizer(c') = \canonizer(c'')$}
\TrinaryInfC{$q_2 \circ q_1: c \Leftrightarrow c'''$}
\end{prooftree}

This derivation rule essentially says that the composition $q_2 \circ q_1$ is
well defined if and only if the intermediary QREs $c''$ and $c'''$ define the
same equivalence relation on the same regular language. The condition
$\mathcal{L}(W(c')) = \mathcal{L}(W(c''))$ says that the intermediary language
is the same on both sides, while the condition $K(c') \equiv K(c'')$ says that
the kernel regular expressions are syntactically the same. Finally, the
condition $\canonizer(c') = \canonizer(c'')$ says that $c'$ and $c''$ define
the same equivalence relation on $\mathcal{L}(W(c'))= \mathcal{L}(W(c''))$.

In general, checking that $\canonizer(c') = \canonizer(c'')$ is an undecidable
task, especially because of the $\normalize{R}{R'}{f}$ QRE. Therefore, our
actual implementation of the composition rule for QRE lenses follows Boomerang's
approach, which is to define the composition on QRE lenses if and only if
$c'$ and $c''$ are syntactially equal to $id(R)$ and $id(R')$ for some $R$and
$R'$ satisfying $\mathcal{L}(R) = \mathcal{L}(R')$.
